Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-12-27 docs/transport/Response_to_Reviewer_1.tex

\documentclass[11pt]{article}

\usepackage{todonotes}
\usepackage{fullpage}
\usepackage{amsmath,amsfonts,amssymb,stmaryrd}
\usepackage{color}
\newcommand{\comment}[1]{{\color{blue} #1}}

\begin{document}

\noindent
Dear Silas Alben,
\\ \\
\noindent
Thank you for handling the manuscript, and we would like to thank the
reviewer for their positive and constructive comments.  The attached
manuscript addresses all the reviewer's comments and the result is what
we believe is a stronger manuscript.  We also note that formatting
issues were resolved in the bibliography. An itemized list of the
changes addressing the reviewer's comments are below.
\\ \\
\noindent
Sincerely,
\\
\noindent
Shang-Huan Chiu, Nick Moore, and Bryan Quaife

\section*{Response to Reviewer 1}

\noindent
\comment{The Barycentric formulae (5.9 \& 5.10) for the first
derivatives of Cauchy integral are incorrect. They are NOT true
barycentric form and straightforward numerical experiments will show
that all accuracies are lost up to the boundary using these formulae,
even if $v(x)$ is computed to high accuracy using the formulae (5.6 \&
5.7). (See e.g.~Sec 3.3 of Barnett et.~al.~2015 for more details.)}
\begin{itemize}
  \item response
\end{itemize}

\noindent
\comment{Likewise, the second derivative ``barycentric formulae" (5.15 \& 5.18) are incorrect, either. So the authors will need to explain how they managed to get high accuracies with the stated formulae?}
\begin{itemize}
  \item response
\end{itemize}

\noindent
\comment{In my opinion, the claim of having developed a ``new quadrature
method" is too strong. After all, there is nothing new about the
barycentric quadrature itself. Rather, I suggest that maybe a more
appropriate claim would be that the authors have developed a ``new
numerical scheme" for the Stokes porous media flow, where the accurate
quadrature is one ingredient in this scheme.}
\begin{itemize}
  \item response
\end{itemize}

\noindent
\comment{The transient increase of tortuosity is interesting, but the
explanation in Sec 6.4 is not satisfying. The authors proposed that the
increase of tortuosity can be explained by the sudden increase of
$x$-velocity when a grain vanishes. But in formula (4.4) or (4.5), the
$x$-velocity appears both in the numerator and the denominator, which
seems to me that the increase would just cancel. Then, as the authors
have pointed out, wouldn't the straighter streamlines resulted from the
vanishing of grains have decreased the tortuosity?}
\begin{itemize}
  \item response
\end{itemize}

\noindent
\comment{The porosity-tortuosity models (4.6c \& 4.6d) are not
explicitly compared in the results section. How much worse are they
comparing to the other models?}
\begin{itemize}
  \item response
\end{itemize}

\noindent
\comment{The coefficient CE in equation (2.1) is undefined.}
\begin{itemize}
  \item Thank you for pointing this out. We also realized that $\mu$ is
  undefined. We now state:
  \begin{quotation}
    \noindent
    Parameters include the fluid viscosity $\mu$ and the
    material-dependent erosion constant $C_E$. By non-dimensionalizing
    equations (2.1) (Quaife \& Moore 2018), we set both these parameters
    to one.
  \end{quotation}
\end{itemize}

\noindent
\comment{Should the shear stress formula (2.2) contain viscosity in it
based on the governing equation (2.1)?}
\begin{itemize}
  \item In general, it would. However, since we have non-dimensionalized
    following our previous work, the viscosity is assumed to be 1.
\end{itemize}

\noindent
\comment{In Figure 5, do the colors represent vorticity itself or the
log of the abs vorticity? (Which would be consistent with the colors in
Figure 4.)}
\begin{itemize}
  \item The color represents the logarithm of the absolute value of the
    vorticity.
  \item We have updated the caption to read:
    \begin{quotation}
    \noindent
      The eroded bodies from figure 4 at time $t = 0.1$. The color is
      the logarithm of the magnitude of the vorticity. The initial
      distances from the body to the solid wall are: (a) $h$, (b) $h/2$,
      and (c) $h/10$.
    \end{quotation}
  \item We have also updated the text to read:
    \begin{quotation}
    \noindent
      To visualize the flow, in figure 5 we plot the logarithm of the
      absolute value of the vorticity from the final time step in figure
      4.
    \end{quotation}
\end{itemize}

\noindent
\comment{When first introduced in Sec 4.3, the ``pore sizes" are not
clearly defined. (My first impression was that pore size = portion of an
edge of the Delaunay diagram. Only at the end in Sec 6.4 do we know that
it is defined as the closest distance between grains.)}
\begin{itemize}
  \item In Figure 3, we clarify the definition by adding the statement:
    {\em represented by the length of the red line segments}.
  \item In the text, we now state:
    \begin{quotation}
    \noindent
      Finally, we define the pore sizes to be the minimum distance
      between all pairs of neighboring grains, and these distances are
      represented by the red line segments in figure 3(a).
    \end{quotation}
\end{itemize}

\noindent
\comment{``Porosity" is undefined throughout the paper.}
\begin{itemize}
  \item We have now formally defined the porosity by rewriting the
    description of coarse-grained models as:
  \begin{quotation}
  \noindent
    Examples of coarse-grained models for porous media flow include
    tortuosities (Matyka et al.~2008; Duda et al.~2011; Koponen et
    al.~1996), geometry connectivity (Knudby \& Carrera 2005), anomalous
    dispersion (Dentz et al.~ 2004), and permeability-porosity
    relationships (Dardis \& McCloskey 1998; Carman 1937), where the
    porosity is the fraction of the geometry that is fluidized.
  \end{quotation}

\end{itemize}

\noindent
\comment{``Area fraction" is plotted in (and only in) Figure 7, while
the rest of the paper used porosity. Maybe plotting porosity here would
give a more consistent and smoother reading experience?}
\begin{itemize}
  \item We examined the area fraction and flow rate in figure 7 so that
    it could be compared to results from our previous work. The goal is
    to demonstrate that we now able to resolve smaller area fractions
    and correspondingly lower flow rates.

  \item We have updated the text to read:
    \begin{quotation}
      \noindent
      In our previous work (Quaife \& Moore 2018), we investigated the
      effects of erosion on the area fraction and flow rate of two
      different geometries. Here, we do the same analysis, but with the
      denser packing of the grains.
    \end{quotation}

\end{itemize}

\end{document}

### 2019-12-27 docs/transport/Response_to_Reviewer_2.tex

\documentclass[11pt]{article}

\usepackage{todonotes}
\usepackage{fullpage}
\usepackage{amsmath,amsfonts,amssymb,stmaryrd}
\usepackage{color}
\newcommand{\comment}[1]{{\color{blue} #1}}

\begin{document}

\noindent
Dear Silas Alben,
\\ \\
\noindent
Thank you for handling the manuscript, and we would like to thank the
reviewer for their positive and constructive comments.  The attached
manuscript addresses all the reviewer's comments and the result is what
we believe is a stronger manuscript.  We also note that formatting
issues were resolved in the bibliography. An itemized list of the
changes addressing the reviewer's comments are below.
\\ \\
\noindent
Sincerely,
\\
\noindent
Shang-Huan Chiu, Nick Moore, and Bryan Quaife

\section*{Response to Reviewer 2}

\noindent
\comment{equations 2.1.  I am confused as to how the no-slip condition
is implemented as zero velocity when in fact the interface is receding
at a rate Vn.}
\begin{itemize}
  \item This comes from a quasi-steady approximation that we have used
    in our previous work as well as by others. For example, see Section
    5.2 from Mitchell and Spagnolie's JCP paper {\em A generalized
    traction integral equation for Stokes flow, with applications to
    near-wall particle mobility and viscous erosion}.

  \item In Section 2, we now state:
    \begin{quotation}
      \noindent
      Since the rate of erosion is much slower than the background flow,
      a quasi-steady approximation justifies the no-slip boundary
      condition.
    \end{quotation}
\end{itemize}

\noindent
\comment{end of section 4.1.  It seems that this protocol (matching
velocity but not position for reinsertion) results in each particle
moving through a different landscape.  Can the authors comment on the
significance of this?}
\begin{itemize}
  \item The reinsertion approach we take has been shown to not introduce
    any significantly artificial decorrelation.

  \item In Section 4.1, We have added that
    \begin{quotation}
      \noindent
      This method of reinsertion does not artificially decorrelate the
      particle velocities (Puyguiraud et al.~2019), and this allows us
      to reliably use the statistical moments (4.2) to characterize the
      anomalous dispersion.
    \end{quotation}

\end{itemize}

\noindent
\comment{p.10, end of section 4.  The authors do not explicitly say how
pore size is calculated.  Is it simply the length of the portion of a
Delaunay side that lives outside the grains?}
\begin{itemize}
 \item In Figure 3, we clarify the definition by adding the statement:
    {\em represented by the length of the red line segments}.
  \item In the text, we now state:
    \begin{quotation}
    \noindent
      Finally, we define the pore sizes to be the minimum distance
      between all pairs of neighboring grains, and these distances are
      represented by the red line segments in figure 3(a).
    \end{quotation}
\end{itemize}

\noindent
\comment{p.17 ``single body close to a wall", the 'experiment' mentioned
here is, I believe, a numerical `experiment'.  It might be good to
indicate that clearly.}
\begin{itemize}
  \item The reviewer is correct that this is a numerical experiment.

  \item There were other instances in the paper where we claimed to be
    performing an `experiment'. We have clarified each of the instances
    by referring them as a `numerical experiment'.
\end{itemize}

\noindent
\comment{sections 6.2 and 6.3.  How were these geometries of 20 grains
chosen?  Both the sizes and positions of these grains vary between
examples - it's not just the porosity that is different.}
\begin{itemize}
  \item The initial placement and radii of the bodies are chosen with a
    randomized method. The algorithm is descried at the start of Section
    6.

  \item We have updated the start of section 6.2 to read:
    \begin{quotation}
      \noindent
      We randomly initialize 20 eroding grains with an initial porosity
      of $\phi = 37.68\%$ and the Hagen-Poiseuille flow
      $\mathbf{U}(\mathbf{x}) = U (1-y^2, 0)$ imposed on $\Gamma$
    \end{quotation}

  \item We have updated the start of section 6.3 to read:
    \begin{quotation}
      \noindent
      We consider a second example with of 20 randomly initialized
      eroding bodies, but with a smaller initial porosity of $\phi =
      30.67\%$.
    \end{quotation}

  \item We have updated the start of section 6.4 to read:
    \begin{quotation}
      \noindent
        As a final example, we consider 100 randomly initialized eroding
        bodies with an initial porosity near $50\%$.
    \end{quotation}

\end{itemize}

\noindent
\comment{Figure 8.  Due to what I presume is an artifact of plotting, it
looks like lots of streamlines suddenly end, which is confusing.}
\begin{itemize}
  \item In figure 8, we freeze an eroded geometry and plot the
    trajectory of passive tracers at five different time steps. Since
    the flow is steady and the tracers are passive, the trajectories
    trace out streamlines.

  \item We have updated the caption to read:
    \begin{quotation}
      \noindent
      200 particle trajectories in the second geometry from figure 6.
      The particles are initially equispaced at $(−1,y)$, where $y \in
      (−1, 1)$. The first five snapshots are equispaced in time. The
      bottom right frame is a magnification of the fifth snapshot, but
      with additional trajectories.
    \end{quotation}

  \item We have also updated the paragraph describing figure 8 to
    clarify that it is showing tracer trajectories that trace out
    streamlines.
\end{itemize}

\noindent
\comment{Figure 9.  While the sharp features in subfigure (b) are
apparently finite cusps, they raise the question of whether any infinite
dwell times (divergent $\tau$) are ever observed in these simulations -
i.e.~are particles ever trapped inside recirculating zones?}
\begin{itemize}
  \item Thank you for pointing out this possibility. The
    recirculation zones are close to the eroding bodies and are
    typically very small. Because we measure the local tortuosity by
    initiating streamlines significantly far from the grains, each of
    them connect the inlet to the outlet.

  \item We have updated the text to read:
    \begin{quotation}
      \noindent
      We note that if a streamline originates in a recirculation zone,
      which is possible in two dimensions (Higdon 1985), then the local
      tortuosity will be infinite. However, in each of our numerical
      experiments, all the streamlines connect the inlet to the outlet.
      Having only finite local tortuosities, the hydraulic tortuosity is
      defined by taking the average over all points on the inlet cross-
      section
    \end{quotation}
\end{itemize}

\noindent
\comment{p.26 bottom paragraph typo: ``an ... media"}
\begin{itemize}
  \item This typo has been fixed
\end{itemize}

\end{document}

### 2019-12-27 docs/transport/Response_to_Reviewer_3.tex

\documentclass[11pt]{article}

\usepackage{todonotes}
\usepackage{fullpage}
\usepackage{amsmath,amsfonts,amssymb,stmaryrd}
\usepackage{color}
\newcommand{\comment}[1]{{\color{blue} #1}}

\begin{document}

\noindent
Dear Silas Alben,
\\ \\
\noindent
Thank you for handling the manuscript, and we would like to thank the
reviewer for their positive and constructive comments.  The attached
manuscript addresses all the reviewer's comments and the result is what
we believe is a stronger manuscript.  We also note that formatting
issues were resolved in the bibliography. An itemized list of the
changes addressing the reviewer's comments are below.
\\ \\
\noindent
Sincerely,
\\
\noindent
Shang-Huan Chiu, Nick Moore, and Bryan Quaife

\section*{Response to Reviewer 3}

\noindent
\comment{Pores in natural porous media care of three-dimensional
geometry.  The authors may include some discussion on how to extend
their numerical method to three-dimensional problems.  In addition, they
may also discuss the limitations of the erosion analysis in
2-dimensional flow to be applied for three-dimensional flow. If the
erosion process in three-dimensional flow is very different from that in
two-dimensional flow, the analysis of this paper may have less relevant
to the analysis of erosion in natural porous media.}
\begin{itemize}
  \item The reviewer is correct that by assuming the geometry is
    two-dimensional, we have limited the applicability for natural
    porous media. There are several numerical challenges to simulate
    three-dimensional erosion, with the most significant being the
    $\theta-L$ formulation, quadrature method, spectral representation
    of the geometry, and the boundary integral equation solver.

  \item Others have developed three-dimensional numerical methods for
    removing stiffness in interfacial flow ($\theta-L$ formulation),
    spectrally representing geometries, and efficiently solving BIEs.
    The challenge with these methods is their computational cost. To
    consider suspensions with the resolution and number of grains that
    we consider, the BIE would most likely require a parallel
    implementation that would run on high-performance computing
    architectures. At this point, this is not the goal of our work.

  \item The Barycentric quadrature method relies on a complex variable
    formulation, so it is inherently two-dimensional. However, there are
    other quadrature methods that can be applied in three dimensions,
    but they are also computationally expensive.

  \item Our response to the reviewer's second comment points to a
    paragraph that has been added to the Introduction.

\end{itemize}

\noindent
\comment{In the numerical examples of the paper the grains are all
separated. In natural porous media, grains are compressed to form grain
contacts. The authors may include discussion on numerical challenges in
studying erosion process in natural porous media.}
\begin{itemize}
  \item The reviewer is correct that porous media grains are compressed
    and form grain contact. This is the result of processes such as
    sedimentation and transport of the grains themselves. This, again,
    is a limitation of our method since we only consider immobile
    bodies.

  \item We plan to consider mobile bodies in the future, and this will
    require a significant computational effort.

  \item To the best of our knowledge, nobody has studied the effect
    erosion has on transport through a porous media. Therefore, as a
    first step, we limit ourselves to two-dimensional suspensions of
    immobile bodies, and we plan to remove these limitations in the
    future. In the Introduction, we have added the following paragraph:
    \begin{quotation}
      \noindent
      While natural porous media are three-dimensional, there are
      several challenges that must be overcome to extend our high-order
      methods to three dimensions.  A three-dimensional version of a
      {$\theta-L$} formulation exists (Ambrose {\em et al.} 2013), but
      the Barycentric quadrature method relies upon a complex variable
      formulation of the BIE, making it inherently two-dimensional.
      Therefore, a different quadrature method is required. Additional
      challenges of three-dimensional simulations include spectral
      representations of the geometry and efficiently solving
      discretizations of the BIE.  While high-order methods for
      three-dimensional BIEs have been developed, their computational
      cost limits the number of bodies that can be practically
      simulated. Another limitation of this work is the assumption of
      immobile grains.  Mechanisms such as sedimentation play a role in
      real applications, and this will effect the transport properties.
      Simulating mobile grains introduces the additional challenge of
      resolving contact.  Since high-order methods have never been used
      to study transport in eroded porous media, as a first step we
      consider the case  with two-dimensional immobile grains. This
      assumption allows for a clear interpretation of the results.  For
      example, in this work we are able to simulate a large
      densely-packed suspension (up to $100$ in this manuscript).

    \end{quotation}
\end{itemize}

\end{document}

### 2019-12-27 docs/proposals/NSF/ProjDesc/ProjDesc.tex
NM
Responses for JFM paper

%---------------------------------------------------------------------------------------------------%
%% TITLE
\begin{document}
\begin{center}
{\large \bf Erosion, Transport, and Dispersion in Granular
and Porous Media} 
\\
M.J.N.~Moore and Bryan D.~Quaife
\end{center}
%\title{Project Description --
%Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media}

@@

The PIs co-authored a paper in the {\em Journal of Computational Physics} describing numerical methods for simulating erosion in the Stokes flow regime~\cite{Quaife2018}, the regime relevant for the currently proposed groundwater applications. Very recently the PIs, along with postdoctoral scholar Shang-Huan Chiu, submitted a second paper on newly developed methods for precise characterization of transport and dispersion within the special class of densly packed porous-media configurations formed by erosion~\cite{chi-moo-qua2019}.

@@

{\bf An immediate goal of the proposed work is to extend the range of
physical effects captured in these simulations.} A particular challenge
of a BIE formulation is developing quadrature for nearly-singular
integrands that arise when bodies are close. To address this problem, we
have very recently extended a Barycentric quadrature
method~\cite{bar2014, bar-wu-vee2015} and have successfully simulated
erosion with grains that are initially separated by 1\% of an arclength
spacing~\cite{chi-moo-qua2019}. The quadrature method is a non-intrusive
modification of the trapezoid rule and results in an error bound that is
independent of the grain configuration. However, the current
implementation does not yet fully exploit FMM acceleration. The
computational challenge is to develop FMM expansions for the velocity,
vorticity, pressure, and deformation-tensor layer potentials. Since these layer potentials each have distinct kernels, we will utilize kernel-independent FMMs~\cite{yin-bir-zor2004, fon-dar2009}, and their HPC implementations (for example, see \url{http://pvfmm.org}). Other challenges to be investigated include preconditioning strategies to reduce the number of required GMRES iterations, and time adaptivity to enable stable simulations to user-specified tolerances. PI Quaife is an expert in developing such algorithms~\cite{qua-bir2015b, qua-bir2016, qua-bir2015a, qua-cou-dar2018}.

>>>>

%---------------------------------------------------------------------------------------------------%
%% TITLE
\begin{document}
\begin{center}
{\large \bf Project Description: \\
Erosion, Transport, and Dispersion in Granular
and Porous Media} 
\\
M.J.N.~Moore and Bryan D.~Quaife
\end{center}
%\title{Project Description --
%Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media}

@@

The PIs co-authored a paper in the {\em Journal of Computational Physics} describing numerical methods for simulating erosion in the Stokes flow regime~\cite{Quaife2018}, the regime relevant for the currently proposed groundwater applications. Very recently the PIs, along with postdoctoral scholar Shang-Huan Chiu, submitted a second paper on newly developed methods for precise characterization of transport and dispersion within the special class of densely packed porous-media configurations formed by erosion~\cite{chi-moo-qua2019}.

@@

{\bf An immediate goal of the proposed work is to extend the range of physical effects captured in these simulations.} A particular challenge of a BIE formulation is developing quadrature for nearly-singular integrands that arise when bodies are close. To address this problem, we have very recently extended a Barycentric quadrature method~\cite{bar2014, bar-wu-vee2015} and have successfully simulated erosion with grains that are initially separated by 1\% of an arclength spacing~\cite{chi-moo-qua2019}. The quadrature method is a non-intrusive modification of the trapezoid rule and results in an error bound that is independent of the grain configuration. However, the current implementation does not yet fully exploit FMM acceleration. The computational challenge is to develop FMM expansions for the velocity, vorticity, pressure, and deformation-tensor layer potentials. Since these layer potentials each have distinct kernels, we will utilize kernel-independent FMMs~\cite{yin-bir-zor2004, fon-dar2009}, and their HPC implementations~\cite{pvfmm}. Other challenges to be investigated include preconditioning strategies to reduce the number of required GMRES iterations, and time adaptivity to enable stable simulations to user-specified tolerances. PI Quaife is an expert in developing such algorithms~\cite{qua-bir2015b, qua-bir2016, qua-bir2015a, qua-cou-dar2018}.
